\chapter{Getting Started} \label{ch:getting_started}

\vspace{1em}
\begin{center}
\parbox{5.5in}{ \emph{
\textbf{I have a multi-objective optimization problem to solve, how
do I use JEGA?}} }
\end{center}
\vspace{1em}

JEGA can be used in a number of ways.  Regardless of how you wish to
use it, there are some common things that must be done.

\section{Obtaining JEGA} \label{sec:obtaining_JEGA}

Currently, JEGA can only be publicly obtained through
\href{http://dakota.sandia.gov}{The DAKOTA Project}.  Downloading
DAKOTA will result in a download of the full JEGA package as well.

\section{Building JEGA}\label{sec:building_JEGA}

JEGA is primarily meant to be compiled into a static library.  It is
intended to be used from within an existing or newly created
program.  There is a way to use JEGA as a stand alone application
and provide input via an input file.  See ?? for more details.
Regardless of how you are using JEGA, the first step is to compile
it. This of course only need be done once and then the JEGA library
can be used in multiple projects. JEGA uses an autoconf build
harness for Unix style systems and Visual Studio .NET 2003, 2005, and
2008 solution and project files for Windows. Build JEGA with your desired
configuration options and make note of the name and location of the
resulting library file.  See Chapter~\ref{ch:configuration} for
information on the configuration options.

\subsection{Building on Unix and Unix-like Systems}\label{sec:building_unix}

The JEGA distribution includes the files produced by autoconf.  The
end user need only issue two commands to compile JEGA.  The first is
a call to the configure script.  You will issue this call with any
desired arguments as described in Section~\ref{sec:compile_config}.
It is strongly recommended that you maintain separate build and
source trees.  In order to accomplish this, you should issue the
configure call from a directory other than \texttt{\$(JEGA\_ROOT)}.
Recommended practice is to create subfolders within the
\texttt{\$(JEGA\_ROOT)/build} directory.  This way, you can maintain
multiple configurations on multiple different platforms/machines,
etc. simultaneously.

Once configuration is complete, JEGA can be built with a call to
make.  The commands that must be issued are shown below.

\texttt{>> \$(JEGA\_ROOT)/configure ARGS}\newline
\texttt{>> make}

As an example, suppose you have multiple machines on which you may
want to use JEGA.  Suppose that the JEGA source distribution is on a
shared drive and you wish to share a source tree but not a build
tree. Also suppose that you wish to maintain both a debug and a
release build.  Here is the recommended procedure. Lets say the two
machines are named yin and yang.

- Browse into \texttt{\$(JEGA\_ROOT)/build}:\newline
    \texttt{>> cd \$(JEGA\_ROOT)/build}\newline

- Make a directory for each of yin and yang:\newline
    \texttt{>> mkdir yin}\newline
    \texttt{>> mkdir yang}\newline

- Make subdirectories in each for debug and release:\newline
    \texttt{>> cd yin}\newline
    \texttt{>> mkdir debug}\newline
    \texttt{>> mkdir release}\newline
    \texttt{>> cd ../yang}\newline
    \texttt{>> mkdir debug}\newline
    \texttt{>> mkdir release}\newline

- Enter the yin debug directory while on yin, configure, and make:\newline
    \texttt{>> <LOGON> yin}\newline
    \texttt{>> cd \$(JEGA\_ROOT)/build/yin/debug}\newline
    \texttt{>> \$(JEGA\_ROOT)/configure CXXFLAGS=-g --enable-debugging}\newline
    \texttt{>> make}\newline

- Enter the yin release directory while on yin, configure, and make:\newline
    \texttt{>> cd ../release}\newline
    \texttt{>> \$(JEGA\_ROOT)/configure CXXFLAGS=-O2}\newline
    \texttt{>> make}\newline

- Do the same in the yang directories while on yang.\newline

Once you have completed these steps, you will have static libraries
in each of the 4 directories built from the same shared source tree.

\subsection{Building in Microsoft Visual Studio}\label{sec:building_msvs}

JEGA is distributed with solution and project files for both Visual
Studio .NET 2003 (7.1), 2005 (8.0), and 2008 (9.0).  The solution files
can be found in \texttt{\$(JEGA\_ROOT)/build/vc71},
\texttt{\$(JEGA\_ROOT)/build/vc80}, \texttt{\$(JEGA\_ROOT)/build/vc90}
respectively.  Open the desired ``.sln'' file.  There are a number of
build configurations available.  Each is set up to build JEGA with
different options. See Table~\ref{} below for a listing of those options.
Choose your desired build configuration and then build the project.  The
result will be a library against which you can link and an executable if
building the configuration file front end.

\subsection{Building in Eclipse Using CDT}\label{sec:building_eclipse_cdt}



\section{Using JEGA Once Built}\label{sec:using_built_JEGA}

In order to get started using the created library, you will have to
create a project. Exactly what this requires depends on what system
you are developing on.  If developing on Window for example, you
will want to create a new Visual Studio project or a new nmake
project. If on a unix type system, you will create a new makefile
project, etc.  If you already have a project that you want to start
using JEGA in, you need only modify it slightly to start using JEGA.

Regardless of what system you are using, there are some common
requirements for using JEGA.

\subsubsection{Required Include Paths}\label{sec:req_incl_paths}

You will have to add some paths to your list of include directories.
This is typically done with a command line argument to your compiler
such as -I or /I.

\begin{itemize}
\item \texttt{\$(JEGA\_ROOT)/include}
\item \texttt{\$(JEGA\_ROOT)/eddy}
\end{itemize}

There may be additional include path requirements depending on what
system you are using.  See the sections below for specific
requirements.

Windows does not come with an implementation of pthreads which JEGA
uses if compiled with \texttt{JEGA\_THREADSAFE} defined. To deal
with this case, JEGA is distributed with a snapshot of the PThreads
Win32 headers and libraries (http://sourceware.org/pthreads-win32/).
If you need to use them, then you must also include the path:

\begin{itemize}
\item \$(JEGA\_ROOT)/eddy/threads/pthreads/include
\end{itemize}

\subsubsection{Required Libraries}\label{sec:req_libs}

You will also have to link against the JEGA library created during
the build. This is typically specified with two inputs to the
compiler. One to specify a path to look in and one to specify the
name of the library. The library created when building JEGA will be
called libjega.\$(LIB\_EXT) where \$(LIB\_EXT) is the file extension
for static libraries on your system; typically .a or .lib.


The easiest thing to do is to follow the examples in
Chapter~\ref{ch:examples}.

\subsection{Using JEGA on Unix and Unix-like Systems}\label{sec:using_on_unix}

If on a unix like system, you will want to add the paths to the
configuration headers that were created by autoconf. There are two
paths needed for two headers. They are in the build tree, not the
source tree. Consider \texttt{\$(BUILD\_DIR)} to be the path to the
trunk of your build directory in the discussion below. In the
example of Section~\ref{sec:building_unix}, \texttt{\$(BUILD\_DIR)}
would be one of \texttt{\$(JEGA\_ROOT)/build/yin/debug},
\texttt{\$(JEGA\_ROOT)/build/yang/release}, etc.  The directories
you will need to add are:

\begin{itemize}
\item \texttt{\$(BUILD\_DIR)}
\item \texttt{\$(BUILD\_DIR)/eddy}
\end{itemize}

\subsection{Using JEGA in Microsoft Visual Studio}\label{sec:building_msvs}
