\chapter{Operators} \label{ch_operators}

The JEGA (John Eddy's Genetic Algorithms) library %[\ref JEddy2001 "Eddy and Lewis, 2001"]
contains two global optimization methods. The first is a
Multi-objective Genetic Algorithm (MOGA) which performs Pareto
optimization.  The second is a Single-objective Genetic Algorithm
(SOGA) which performs optimization on a single objective function.
Both methods support general constraints and a mixture of real and
discrete variables.  The JEGA library was written by John Eddy,
currently a member of the technical staff in the System Sustainment
and Readiness group at Sandia National Laboratories in Albuquerque.
These algorithms are accessed as \emph{moga} and \emph{soga} within
DAKOTA.  DAKOTA provides access to the JEGA library through the
JEGAOptimizer class.

JEGA utilizes the \emph{max\_iterations} and
\emph{max\_function\_evaluations} method independent controls to
provide integer limits for the maximum number of generations and
function evaluations, respectively.  Note that currently, the DAKOTA
default for \emph{max\_iterations} is 100 and for
\emph{max\_function\_evaluations} is 1000. These are the default
settings that will be used to ``stop'' the JEGA algorithms, unless
some specific convergence criteria are set% (see \ref T5d20 "Table
%5.20" below).

JEGA v2.0 also utilizes the \emph{output} method independent control to
vary the amount of information presented to the user during
execution.

%\subsection MethodJEGADC JEGA method dependent controls

The JEGA library currently provides two types of genetic algorithms
(GAs): a multi-objective genetic algorithm (\emph{moga}), and a
single- objective genetic algorithm (\emph{soga}).  Both of these
GAs can take real-valued inputs, integer-valued inputs, or a mixture
of real and integer-valued inputs.  ``Real-valued'' and
``integer-valued'' refer to the use of continuous or discrete
variable domains, respectively (the response data are real-valued in
all cases).

The basic steps of the genetic algorithm are as follows:

\begin{itemize}

\item Initialize the population.  This is typically done by randomly
generating population members with or without duplicates or by some
sort of ``warm start'' initialization.

\item Evaluate the initial population members.  This involves the
calculation of the values of the objective function(s) and
constraints for each population member.

\item Perform crossover.  This is where existing members of the
population are combined to form new ones.  Several crossover types
are available.  This is the beginning of the iterator loop.

\item Perform mutation.  This is where random variation is added
to the offspring designs.  Several mutation types are available.

\item Evaluate the new population members.

\item Assess the fitness of each member in the population.  There are
a number of ways to evaluate the fitness of the members of the
populations.  Choice of fitness assessor operators is strongly
dependent on the type of algorithm being used and can have a
profound effect on the choice of selectors.  For example, if using
\emph{MOGA}, the available assessors are the \emph{layer\_rank} and
\emph{domination\_count} fitness assessors.  If using either of
these, it is strongly recommended that you use the
\emph{below\_limit} selector as well (although the roulette wheel
selectors can also be used).  The functionality of the
\emph{dominationr\_count} selector of JEGA v1.0 can now be achieved
using the \emph{domination\_count} fitness assessor and
\emph{below\_limit} selector together.  If using \emph{SOGA}, the
only fitness assessor is the \emph{merit\_function} fitness assessor
which currently uses an exterior penalty function formulation to
assign fitnesses. Any of the selectors can be used in conjunction
with this fitness assessment scheme.  This is the first of the
selection operators.

\item Replace the population with members selected to continue in the
next generation.  The pool of potential members is the current
population and the current set of offspring.  The
\emph{replacement\_type} of \emph{roulette\_wheel} or
\emph{unique\_roulette\_wheel} may be used either with MOGA or SOGA
problems however they are not recommended for use with MOGA.  Given
that the only two fitness assessors for MOGA are the
\emph{layer\_rank} and \emph{domination\_count}, the recommended
selector is the \emph{below\_limit} selector. The
\emph{replacement\_type} of \emph{favor\_feasible} is specific to a
SOGA. This replacement operator will always take a feasible design
over an infeasible one.  Beyond that, it favors solutions based on
an assigned fitness value which must have been installed by some
fitness assessor.

\item Apply niche pressure to the population.  This step is specific
to the MOGA and is new to JEGA v2.0.  Technically, the step is
carried out during runs of the SOGA but only the
\emph{null\_niching} operator is available for use with SOGA.  In
MOGA, the \emph{radial} niching operator or the \emph{distance}
niching operator can be used. The purpose of niching is to encourage
differentiation along the Pareto frontier and thus a more even and
uniform sampling.  In JEGA, niching operators typically work by
determining when two designs are too close in the performance
(phenotype) space and caching them until the next round of
selection.

\item Test for convergence.  The final step in the iterator loop is
to assess the convergence of the algorithm.  There are two aspects
to convergence that must be considered.  The first is stopping
criteria.  A stopping criteria dictates some sort of limit on the
algorithm that is independent of its performance.  Examples of
stopping criteria available for use with JEGA are the
\emph{max\_iterations} and \emph{max\_function\_evaluations} inputs.
All JEGA convergers respect these stopping criteria in addition to
anything else that they do.

The second aspect to convergence involves repeated assessment of the
algorithms progress in solving the problem.  In JEGA v1.0, the
fitness tracker convergers (\emph{best\_fitness\_tracker} and
\emph{average\_fitness\_tracker}) performed this function by
asserting that the fitness values (either best or average) of the
population continue to improve.  There was no such operator for the
MOGA.  As of JEGA v2.0, the same fitness tracker convergers exist
for use with SOGA and there is now a converger available for use
with the MOGA. The MOGA converger (\emph{metric\_tracker}) operates
by tracking various changes in the non-dominated frontier from
generation to generation. When the changes occurring over a user
specified number of generations fall below a user specified
threshold, the algorithm stops.

\item Post Process.  This occurs after convergence has been
attained.  This operator provides a means of filtering the data
returned by the algorithm in whatever way is relevant.  For example,
if using a MOGA, it may be desired to return only a certain number
of points within a particular region or evenly dispersed throughout
the performance space, etc.
\end{itemize}

There are many controls which can be used for both MOGA and SOGA
methods.  These include among others the random seed, initialization
types, crossover and mutation types, and some replacement types.

The \emph{seed} control defines the starting seed for the random
number generator.  The algorithm uses random numbers heavily but a
specification of a random seed will cause the algorithm to run
identically from one trial to the next so long as all other input
specifications remain the same.  New to JEGA v2.0 is the
introduction of the \emph{log\_file} specification.  JEGA v2.0 uses
a logging library to output messages and status to the user.  JEGA
can be configured at build time to log to both standard error and a
text file, one or the other, or neither.  The \emph{log\_file} input
is a string name of a file into which to log.  If the build was
configured without file logging in JEGA, this input is ignored.  If
file logging is enabled and no \emph{log\_file} is specified, the
default file name if JEGAGlobal.log is used.  Also new to JEGA v2.0
is the introduction of the \emph{print\_each\_pop} specification. It
serves as a flag and if supplied, the population at each generation
will be printed to a file named ``population<GEN\#>.dat'' where
<GEN\#> is the number of the current generation.

The \emph{initialization\_type} defines the type of initialization
for the GA.  There are three types: \emph{simple\_random},
\emph{unique\_random}, and \emph{flat\_file}.  \emph{simple\_random}
creates initial solutions with random variable values according to a
uniform random number distribution. It gives no consideration to any
previously generated designs.  The number of designs is specified by
the \emph{population\_size}. \emph{unique\_random} is the same as
\emph{simple\_random}, except that when a new solution is generated,
it is checked against the rest of the solutions.  If it duplicates
any of them, it is rejected.  \emph{flat\_file} allows the initial
population to be read from a flat file.  If \emph{flat\_file} is
specified, a file name must be given.

Variables can be delimited in the flat file in any way you see fit
with a few exceptions.  The delimiter must be the same on any given
line of input with the exception of leading and trailing whitespace.
So a line could look like: 1.1, 2.2  ,3.3 for example but could not
look like: 1.1, 2.2  3.3.  The delimiter can vary from line to line
within the file which can be useful if data from multiple sources is
pasted into the same input file. The delimiter can be any string
that does not contain any of the characters .+-dDeE or any of the
digits 0-9.  The input will be read until the end of the file.  The
algorithm will discard any configurations for which it was unable to
retrieve at least the number of design variables.  The objective and
constraint entries are not required but if ALL are present, they
will be recorded and the design will be tagged as evaluated so that
evaluators may choose not to re-evaluate them.  Setting the size for
this initializer has the effect of requiring a minimum number of
designs to create.  If this minimum number has not been created once
the files are all read, the rest are created using the
\emph{unique\_random} initializer and then the \emph{simple\_random}
initializer if necessary.

Note that the \emph{population\_size} only sets the size of the
initial population.  The population size may vary in the JEGA
methods according to the type of operators chosen for a particular
optimization run.

There are many crossover types available.
\emph{multi\_point\_binary} crossover requires an integer number, N,
of crossover points.  This crossover type performs a bit switching
crossover at N crossover points in the binary encoded genome of two
designs.  Thus, crossover may occur at any point along a solution
chromosome (in the middle of a gene representing a design variable,
for example). \emph{multi\_point\_parameterized\_binary} crossover
is similar in that it performs a bit switching crossover routine at
N crossover points. However, this crossover type performs crossover
on each design variable individually. So the individual chromosomes
are crossed at N locations. \emph{multi\_point\_real} crossover
performs a variable switching crossover routing at N crossover
points in the real real valued genome of two designs. In this
scheme, crossover only occurs between design variables
(chromosomes).  Note that the standard solution chromosome
representation in the JEGA algorithm is real encoded and can handle
integer or real design variables.  For any crossover types that use
a binary representation, real variables are converted to long
integers by multiplying the real number by $10^6$ and then
truncating. Note that this assumes a precision of only six decimal
places. Discrete variables are represented as integers (indices
within a list of possible values) within the algorithm and thus
require no special treatment by the binary operators.

The final crossover type is \emph{shuffle\_random}.  This crossover
type performs crossover by choosing design variables at random from
a specified number of parents enough times that the requested number
of children are produced.  For example, consider the case of 3
parents producing 2 children.  This operator would go through and
for each design variable, select one of the parents as the donor for
the child. So it creates a random shuffle of the parent design
variable values. The relative numbers of children and parents are
controllable to allow for as much mixing as desired.  The more
parents involved, the less likely that the children will wind up
exact duplicates of the parents.

All crossover types take a \emph{crossover\_rate}.  The crossover
rate is used to calculate the number of crossover operations that
take place. The number of crossovers is equal to the rate *
\emph{population\_size}.

There are five mutation types allowed.  \emph{replace\_uniform}
introduces random variation by first randomly choosing a design
variable of a randomly selected design and reassigning it to a
random valid value for that variable.  No consideration of the
current value is given when determining the new value.  All mutation
types have a \emph{mutation\_rate}.  The number of mutations for the
\emph{replace\_uniform} mutator is the product of the
\emph{mutation\_rate} and the \emph{population\_size}.

The \emph{bit\_random} mutator introduces random variation by first
converting a randomly chosen variable of a randomly chosen design
into a binary string.  It then flips a randomly chosen bit in the
string from a 1 to a 0 or visa versa. In this mutation scheme, the
resulting value has more probability of being similar to the
original value.  The number of mutations performed is the product of
the \emph{mutation\_rate}, the number of design variables, and the
\emph{population\_size}.

The offset mutators all act by adding an ``offset'' random amount to
a variable value.  The random amount has a mean of zero in all
cases. The \emph{offset\_normal} mutator introduces random variation
by adding a Gaussian random amount to a variable value.  The random
amount has a standard deviation dependent on the
\emph{mutation\_scale}.  The \emph{mutation\_scale} is a fraction in
the range [0, 1] and is meant to help control the amount of
variation that takes place when a variable is mutated.
\emph{mutation\_scale} is multiplied by the range of the variable
being mutated to serve as standard deviation. \emph{offset\_cauchy}
is similar to \emph{offset\_normal}, except that a Cauchy random
variable is added to the variable being mutated.  The
\emph{mutation\_scale} also defines the standard deviation for this
mutator. Finally, \emph{offset\_uniform} adds a uniform random
amount to the variable value.  For the \emph{offset\_uniform}
mutator, the \emph{mutation\_scale} is interpreted as a fraction of
the total range of the variable.  The range of possible deviation
amounts is +/- 1/2 * (\emph{mutation\_scale} * variable range).  The
number of mutations for all offset mutators is defined as the
product of \emph{mutation\_rate} and \emph{population\_size}.

As of JEGA v2.0, all replacement types are common to both MOGA and
SOGA. They include the \emph{roulette\_wheel},
\emph{unique\_roulette\_wheel}, and \emph{below\_limit} selectors.
In \emph{roulette\_wheel replacement}, each design is conceptually
allotted a portion of a wheel proportional to its fitness relative
to the fitnesses of the other Designs.  Then, portions of the wheel
are chosen at random and the design occupying those portions are
duplicated into the next population.  Those Designs allotted larger
portions of the wheel are more likely to be selected (potentially
many times). \emph{unique\_roulette\_wheel} replacement is the same
as \emph{roulette\_wheel} replacement, with the exception that a
design may only be selected once.  The \emph{below\_limit} selector
attempts to keep all designs for which the negated fitness is below
a certain limit.  The values are negated to keep with the convention
that higher fitness is better. The inputs to the \emph{below\_limit}
selector are the limit as a real value, and a
\emph{shrinkage\_percentage} as a real value.  The
\emph{shrinkage\_percentage} defines the minimum amount of
selections that will take place if enough designs are available.  It
is interpreted as a percentage of the population size that must go
on to the subsequent generation.  To enforce this,
\emph{below\_limit} makes all the selections it would make anyway
and if that is not enough, it takes the remaining that it needs from
the best of what is left (effectively raising its limit as far as it
must to get the minimum number of selections).  It continues until
it has made enough selections.  The \emph{shrinkage\_percentage} is
designed to prevent extreme decreases in the population size at any
given generation, and thus prevent a big loss of genetic diversity
in a very short time.  Without a shrinkage limit, a small group of
``super'' designs may appear and quickly cull the population down to
a size on the order of the limiting value.  In this case, all the
diversity of the population is lost and it is expensive to
re-diversify and spread the population.
%
%\anchor T5d17 <table> <caption align = "top"> \htmlonly Table 5.17
%\endhtmlonly
%Specification detail for JEGA method dependent controls: seed,
%output, initialization, mutation, and replacement </caption> <tr>
%<td><b>Description</b> <td><b>Keyword</b> <td><b>Associated Data</b>
%<td><b>Status</b> <td><b>Default</b> <tr> <td>GA Method
%<td>\emph{moga}| \emph{soga} <td>none <td>Required group <td>N/A
%<tr> <td>Random Seed <td>\emph{seed} <td>integer <td>Optional
%<td>Time based seed <tr> <td>Log File <td>\emph{log\_file} <td>string
%<td>Optional <td>JEGAGlobal.log <tr> <td>Population Output
%<td>\emph{print\_each\_pop} <td>none <td>Optional <td>N/A <tr>
%<td>Output verbosity <td>\emph{output} <td>\emph{silent} |
%\emph{quiet} | \emph{verbose} | \emph{debug} <td>Optional
%<td>\emph{normal} <tr> <td>Initialization type
%<td>\emph{initialization\_type} <td>\emph{simple\_random} |
%\emph{unique\_random} | \emph{flat\_file} <td>Required
%<td>unique\_random <tr> <td>Mutation type <td>\emph{mutation\_type}
%<td>\emph{replace\_uniform} | \emph{bit\_random} |
%\emph{offset\_cauchy} | \emph{offset\_uniform} | \emph{offset\_normal}
%<td>Optional group <td>None <tr> <td>Mutation scale
%<td>\emph{mutation\_scale} <td>real <td>Optional <td>0.15 <tr>
%<td>Mutation rate <td>\emph{mutation\_rate} <td>real <td>Optional
%<td>0.08 <tr> <td>Replacement type <td>\emph{replacement\_type}
%<td>\emph{below\_limit} | \emph{roulette\_wheel} |
%\emph{unique\_roulette\_wheel} <td>Required group <td>None <tr>
%<td>Below limit selection <td>\emph{below\_limit} <td>real
%<td>Optional <td>6 <tr> <td>Shrinkage percentage in below limit
%selection <td>\emph{shrinkage\_percentage} <td>real <td>Optional<BR>
%<td>0.9 </table>


%
%</table>
%
%\anchor T5d18 <table> <caption align = "top"> \htmlonly Table 5.18
%\endhtmlonly
%Specification detail for JEGA method dependent controls: crossover
%</caption> <tr> <td><b>Description</b> <td><b>Keyword</b>
%<td><b>Associated Data</b> <td><b>Status</b> <td><b>Default</b> <tr>
%<td>Crossover type <td>\emph{crossover\_type} <td>\emph{multi\_point\_binary} |
%\emph{multi\_point\_parameterized\_binary} |
%    \emph{multi\_point\_real} | \emph{shuffle\_random}
%<td>Optional group <td>none <tr> <td>Multi point binary crossover
%<td>\emph{multi\_point\_binary} <td>integer <td>Required <td>N/A <tr>
%<td>Multi point parameterized binary crossover <td>\c
%multi\_point\_parameterized\_binary <td>integer <td>Required <td>N/A
%<tr> <td>Multi point real crossover <td>\emph{multi\_point\_real}
%<td>integer <td>Required <td>N/A <tr> <td>Random shuffle crossover
%<td>\emph{shuffle\_random} <td>\emph{num\_parents}, \emph{num\_offspring}
%<td>Required <td>N/A <tr> <td>Number of parents in random shuffle
%crossover <td>\emph{num\_parents} <td>integer <td>optional <td>2 <tr>
%<td>Number of offspring in random shuffle crossover <td>\c
%num\_offspring <td>integer <td>optional <td>2 <tr> <td>Crossover rate
%<td>\emph{crossover\_rate} <td>real <td>optional (applies to all
%crossover types) <td>0.8 </table>

%\subsection MethodJEGAMOGA Multi-objective Evolutionary Algorithms

The specification for controls specific to Multi-objective
Evolutionary algorithms are described here.  These controls will be
appropriate to use if the user has specified \emph{moga} as the method.

The initialization, crossover, and mutation controls were all
described in the preceding section.  There are no MOGA specific
aspects to these controls.  The \emph{fitness\_type} for a MOGA may
be \emph{domination\_count} or \emph{layer\_rank}.  Both have been
specifically designed to avoid problems with aggregating and scaling
objective function values and transforming them into a single
objective. Instead, the \emph{domination\_count} fitness assessor
works by ordering population members by the negative of the number
of designs that dominate them.  The values are negated in keeping
with the convention that higher fitness is better.  The
\emph{layer\_rank} fitness assessor works by assigning all
non-dominated designs a layer of 0, then from what remains,
assigning all the non-dominated a layer of -1, and so on until all
designs have been assigned a layer.  Again, the values are negated
for the higher-is-better fitness convention. Use of the
\emph{below\_limit} selector with the \emph{domination\_count}
fitness assessor has the effect of keeping all designs that are
dominated by fewer then a limiting number of other designs subject
to the shrinkage limit.  Using it with the \emph{layer\_rank}
fitness assessor has the effect of keeping all those designs whose
layer is below a certain threshold again subject to a minimum.

New to JEGA v2.0 is the introduction of niche pressure operators.
These operators are meant primarily for use with the moga.  The job
of a niche pressure operator is to encourage diversity along the
Pareto frontier as the algorithm runs.  This is typically
accomplished by discouraging clustering of design points in the
performance space.  In JEGA, the application of niche pressure
occurs as a secondary selection operation.  The nicher is given a
chance to perform a pre-selection operation prior to the operation
of the selection (replacement) operator, and is then called to
perform niching on the set of designs that were selected by the
selection operator.

Currently, the only niche pressure operator available is the
\emph{radial} nicher. This niche pressure applicator works by
enforcing a minimum distance between designs in the performance
space at each generation.  The algorithm proceeds by starting at the
(or one of the) extreme designs along objective dimension 0 and
marching through the population removing all designs that are too
close to the current design.  One exception to the rule is that the
algorithm will never remove an extreme design which is defined as a
design that is maximal or minimal in all but 1 objective dimension
(for a classical 2 objective problem, the extreme designs are those
at the tips of the non-dominated frontier).

The designs that are removed by the nicher are not discarded.  They
are buffered and re-inserted into the population during the next
pre-selection operation.  This way, the selector is still the only
operator that discards designs and the algorithm will not waste time
``re-filling'' gaps created by the nicher.

The niche pressure control consists of two options.  The first is
the \emph{null\_niching} option which specifies that no niche
pressure is to be applied. The second is the \emph{radial\_niching}
option which specifies that the radial niching algorithm is to be
use.  The radial nicher requires as input a vector of fractions with
length equal to the number of objectives.  The elements of the
vector are interpreted as percentages of the non-dominated range for
each objective defining a minimum distance to all other designs. All
values should be in the range (0, 1).  The minimum allowable
distance between any two designs in the performance space is the
euclidian distance defined by these percentages.

Also new to JEGA v2.0 is the introduction of the MOGA specific
\emph{metric\_tracker} converger.  This converger is conceptually
similar to the best and average fitness tracker convergers in that
it tracks the progress of the population over a certain number of
generations and stops when the progress falls below a certain
threshold.  The implementation is quite different however.  The
\emph{metric\_tracker} converger tracks 3 metrics specific to the
non-dominated frontier from generation to generation.  All 3 of
these metrics are computed as percent changes between the
generations.  In order to compute these metrics, the converger
stores a duplicate of the non-dominated frontier at each generation
for comparison to the non-dominated frontier of the next generation.

The first metric is one that indicates how the expanse of the
frontier is changing.  The expanse along a given objective is
defined by the range of values existing within the non-dominated
set.  The expansion metric is computed by tracking the extremes of
the non-dominated frontier from one generation to the next.  Any
movement of the extreme values is noticed and the maximum percentage
movement is computed as:
%
%\begin{equation}
%    Em = \max_{j=1)^{nof} |(range(j, i) - range(j, i-1)) / range(j, i-1)|
%\end{equation}

where Em is the max expansion metric, j is the objective function
index, i is the current generation number, and \emph{nof} is the
total number of objectives.  The range is the difference between the
largest value along an objective and the smallest when considering
only non-dominated designs.

The second metric monitors changes in the density of the
non-dominated set.  The density metric is computed as the number of
non-dominated points divided by the hypervolume of the non-dominated
region of space.  Therefore, changes in the density can be caused by
changes in the number of non-dominated points or by changes in size
of the non-dominated space or both.  The size of the non-dominated
space is computed as:

%\begin{equation}
%    Vps(i) = \prod_{j=1)^{nof} range(j, i)
%\end{equation}

where Vps(i) is the hypervolume of the non-dominated space at
generation i and all other terms have the same meanings as above.

The density of the a given non-dominated space is then:

\begin{equation}
    Dps(i) = Pct(i) / Vps(i)
\end{equation}

where Pct(i) is the number of points on the non-dominated frontier
at generation i.

The percentage increase in density of the frontier is then
calculated as:

\begin{equation}
    Cd = |(Dps(i) - Dps(i-1)) / Dps(i-1)|
\end{equation}

where Cd is the change in density metric.

The final metric is one that monitors the ``goodness'' of the
non-dominated frontier.  This metric is computed by considering each
design in the previous population and determining if it is dominated
by any designs in the current population.  All that are determined
to be dominated are counted. The metric is the ratio of the number
that are dominated to the total number that exist in the previous
population.

As mentioned above, each of these metrics is a percentage.  The
tracker records the largest of these three at each generation.  Once
the recorded percentage is below the supplied percent change for the
supplied number of generations consecutively, the algorithm is
converged.

The specification for convergence in a MOGA can either be
\emph{metric\_tracker} or can be omitted all together.  If omitted,
no convergence algorithm will be used and the algorithm will rely on
stopping criteria only.  If \emph{metric\_tracker} is specified,
then a \emph{percent\_change} and \emph{num\_generations} must be
supplied as with the other metric tracker convergers (average and
best fitness trackers).  The \emph{percent\_change} is the threshold
beneath which convergence is attained whereby it is compared to the
metric value computed as described above.  The
\emph{num\_generations} is the number of generations over which the
metric value should be tracked. Convergence will be attained if the
recorded metric is below \emph{percent\_change} for
\emph{num\_generations} consecutive generations.

The MOGA specific controls are described in %\ref T5d19 "Table 5.19"
below. Note that MOGA and SOGA create additional output files during
execution.  ``finaldata.dat'' is a file that holds the Pareto
members of the population in the final generation.  ``discards.dat''
holds solutions that were discarded from the population during the
course of evolution. It can often be useful to plot objective
function values from these files to visually see the Pareto front
and ensure that finaldata.dat solutions dominate discards.dat
solutions.  The solutions are written to these output files in the
format ``Input1...InputN..Output1...OutputM''. If MOGA is used in a
multi-level optimization strategy (which requires one optimal
solution from each individual optimization method to be passed to
the subsequent optimization method as its starting point), the
solution in the Pareto set closest to the ``utopia'' point is given
as the best solution. This solution is also reported in the DAKOTA
output. This ``best'' solution in the Pareto set has minimum
distance from the utopia point.  The utopia point is defined as the
point of extreme (best) values for each objective function.  For
example, if the Pareto front is bounded by (1,100) and (90,2), then
(1,2) is the utopia point.  There will be a point in the Pareto set
that has minimum L2-norm distance to this point, for example (10,10)
may be such a point.  In SOGA, the solution that minimizes the
single objective function is returned as the best solution.

%
%\anchor T5d19 <table> <caption align = "top"> \htmlonly Table 5.19
%\endhtmlonly
%Specification detail for MOGA method controls </caption> <tr>
%<td><b>Description</b> <td><b>Keyword</b> <td><b>Associated Data</b>
%<td><b>Status</b> <td><b>Default</b> <tr> <td>Fitness type
%<td>\emph{fitness\_type} <td>\emph{layer\_rank} |
%\emph{domination\_count} <td>Required group <td> None <tr> <td>Niche
%pressure type <td>\emph{niching\_type} <td>\emph{radial\_niching}
%<td>Optional group <td> No niche pressure <tr> <td>Radial niching
%<td>\emph{radial\_niching} <td>list of real <td>Optional <td> 0.01
%for all objectives <tr> <td>Convergence type
%<td>\emph{metric\_tracker} <td>none <td>Optional group <td>Stopping
%criteria only <tr> <td>Percent change limit for metric\_tracker
%converger <td>\emph{percent\_change} <td>real <td>Optional <td>0.1
%<tr> <td>Number generations for metric\_tracker converger
%<td>\emph{num\_generations} <td>integer <td>Optional <td>10 </table>


%\subsection MethodJEGASOGA Single-objective Evolutionary Algorithms

The specification for controls specific to Single-objective
Evolutionary algorithms are described here.  These controls will be
appropriate to use if the user has specified \emph{soga} as the method.

The initialization, crossover, and mutation controls were all
described above.  There are no SOGA specific aspects to these
controls.  The \emph{replacement\_type} for a SOGA may be
\emph{roulette\_wheel}, \emph{unique\_roulette\_wheel}, or
\emph{favor\_feasible}.  The \emph{favor\_feasible} replacement type
always takes a feasible design over an infeasible one. Beyond that,
it selects designs based on a fitness value.  As of JEGA v2.0, the
fitness assessment operator must be specified with SOGA although the
\emph{merit\_function} is currently the only one.  The roulette
wheel selectors no longer assume a fitness function.  The
\emph{merit\_function} fitness assessor uses an exterior penalty
function formulation to penalize infeasible designs.  The
specification allows the input of a \emph{constraint\_penalty} which
is the multiplier to use on the constraint violations.

The SOGA controls allow two additional convergence types.  The
\emph{convergence\_type} called \emph{average\_fitness\_tracker}
keeps track of the average fitness in a population.  If this average
fitness does not change more than \emph{percent\_change} over some
number of generations, \emph{num\_generations}, then the solution is
reported as converged and the algorithm terminates. The
\emph{best\_fitness\_tracker} works in a similar manner, only it
tracks the best fitness in the population. Convergence occurs after
\emph{num\_generations} has passed and there has been less than
\emph{percent\_change} in the best fitness value.  Both also respect
the stopping criteria.

The SOGA specific controls are described in %\ref T5d20 "Table 5.20"
below.
%
%\anchor T5d20 <table> <caption align = "top"> \htmlonly Table 5.20
%\endhtmlonly
%Specification detail for SOGA method controls </caption> <tr>
%<td><b>Description</b> <td><b>Keyword</b> <td><b>Associated Data</b>
%<td><b>Status</b> <td><b>Default</b> <tr> <td>Fitness type
%<td>\emph{fitness\_type} <td>\emph{merit\_function} <td>Optional group
%<td> merit\_function <tr> <td>Merit function fitness
%<td>\emph{merit\_function} <td>\emph{real} <td>Optional <td>1.0 <tr>
%<td>Replacement type <td>\emph{replacement\_type}
%<td>\emph{favor\_feasible} | \emph{unique\_roulette\_wheel} |
%\emph{roulette\_wheel} <td>Optional group <td> <tr> <td>Convergence
%type <td>\emph{convergence\_type} <td>\emph{best\_fitness\_tracker} |
%\emph{average\_fitness\_tracker} <td>Optional <td> <tr> <td>Number of
%generations (for convergence test) <td>\emph{num\_generations}
%<td>integer <td>Optional <td>10 <tr> <td>Percent change in fitness
%<td>\emph{percent\_change} <td>real <td>Optional <td>0.1 </table>
